%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sheaf representations and duality in logic
% Steve Awodey
% June 2016
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[12pt]{article}

\usepackage{subfiles}


\input{lambek_preamble.tex}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{
Sheaf Representations and Duality in Logic
}
\author{
Steve Awodey
}
\date{
\today}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Outline}


\begin{enumerate}
\item Gelfand Duality

\item Grothendieck duality for rings

\item Lambek-Moerdijk sheaf representation for toposes

\item Local sheaf representation for toposes

\item Stone duality for Boolean algebras

\item Lawvere duality and others

\item Sheaf representation for pretoposes

\item Stone-type dualities for pretoposes

\end{enumerate}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\subfile{lambek_intro.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subfile{lambek_ch1.tex}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subfile{lambek_ch2.tex}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\subfile{lambek_ch3.tex}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{References}

\begin{itemize}
\item Lambek, J.\ and I.~Moerdijk,\\
 Two sheaf representations of elementary toposes, in 
 A.S.\ Troelstra and D.\ van Dalen (Eds.), \emph{Brouwer Centenary Symposium}, North-Holland, Amsterdam, 1982.
 
\item Lambek, J., On the sheaf of possible worlds,\\
 in: Adamek, Mac Lane (Eds.), \emph{Categorical Topology}, World Scientific, Singapore, 1989.
 
 \item Lambek, J., What is the world of mathematics? 
 Annals of Pure and Applied Logic
Volume 126, Issues 1Ð3, April 2004, Pages 149-158

 http://www.math.mcgill.ca/barr/lambek/pdffiles/WorldMath.pdf
 
\item Awodey, S., Sheaf representation for topoi,\\
 \emph{J.\ Pure Appl.\ Algebra}, 145, pp.~107--121, 2000.
\item Lawvere, F.W., Functorial semantics of algebraic theories,  \emph{Proc.\ Nat.\ Acad.\ Sci.\ USA} 50, pp.~869--72, 1963.
\item Makkai, M., Stone duality for first order logic,\\
 \emph{Adv.\ Math.} 65(2), pp.~97--170,  1987.
\item Awodey, S. and H.~Forssell, First-order logical duality,\\
 \emph{Ann.\ Pure and Appl.\ Logic}, 164(3), pp.~319--348, 2013.
\item Breiner, S., \emph{Scheme representation for first-order logic},\\
 Ph.D. thesis, CMU, 2013.
\end{itemize}

%\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%








%%
\end{document}
%%
